mathml: Translate R expressions to MathML and LaTeX

### 2023-01-28 rj/mathml.Rmd
new organization of subscripts
decorations

---
title: "mathml: Translate R expressions to MathML and LaTeX"
date: "2023-01-21"
abstract: >
  This R\ package translates R\ objects to suitable elements in MathML or LaTeX,
  thereby allowing for a pretty mathematical representation of R\ objects and
  functions in data analyses, scientific reports and interactive web content. In
  the RMarkdown document rendering language, R\ code and mathematical content
  already exist side-by-side. The present package allows to use the same
  R\ objects for both data analysis and typesetting in documents or web content.
  This tightens the link between the statistical analysis and its verbal
  description or symbolic representation, which is another step towards
  reproducible science. User-defined hooks allow to extend the package by
  mapping specific variables or functions to new MathML and LaTeX entities. A
  number of examples and a case study illustrate the use of the package in a
  scientific report.
draft: true
author:  
  - name: Matthias Gondan
    affiliation: Universität Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/gondan-rochon/index.html.en
    orcid: 0000-0001-9974-0057
    email:  Matthias.Gondan-Rochon@uibk.ac.at
  - name: Irene Alfarone
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/alfarone/index.html.en
    email: Irene.Alfarone@uibk.ac.at
    orcid: 0000-0002-8409-8900
    affiliation: Universität Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
type: package
output: 
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: bibliography.bib
preamble:
- \usepackage{cancel}
- \usepackage{amsmath}
Vignette: >
  %\VignetteIndexEntry{mathml: Translate R expressions to MathML and LaTeX}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# To do
#
# paste(x, y, z)--> z, y, z (Covered by quote ())

# Spacing: operator for separating elements in expression. E.g.,
## x ~ y --> x  y ;
## x ~ ~ y --> x    y

# Ellipsis: ldots or cdots. E.g., 
## x[1] + ... + x[n]

# Radicals sqrt(x, y) --> y√x

# Chains: a < b < c

# Sets: some symbols in set relations are missing:
## supset, supseteq, notsubset, subset, subseteq, notin

# Phantom: command that produces a space in place of an argument (partially covered by omit() (?)). E.g.:
## x + phantom(0) + y --> x +    + y
## x + over(1, vphantom(0)) --> x + 1/

# Typeface: plain, bold, italic, bolditalic

# Stylechanges: textstyle(x), scriptstyle(x) scriptscript(x)

# linear models: DepVar ~ X0 + X1

# hat, overline, cancel, box, invisible, phantom

@@

```{r, results='asis'}
term <- quote(1 + -2L + a + abc + "a" + phi + Phi + varphi + hat(b)[i, j]^2L)
mathout(term)

@@

Note that an R\ function `hat()` does not exist in base R, it is provided by the
package for convenience and points to the identity function. Quite a few such
convenience functions have been defined
in \pkg{mathml} (see, e.g., next section, and the manual for details).

@@

Indices in square brackets are rendered as subscripts, powers are rendered as
superscript. Moreover, \pkg{mathml} defines the
functions `subscript()`, `superscript()` and `subsuperscript()` that simply
return their first argument. The other arguments serve as decorations, for
example, for summation and product signs.

@@

term <- quote(log(subsupscript(prod(L[i]), i==1L, N)) <- 
          subsupscript(sum(log(L[i])), i==1L, N))
mathout(term)
```

@@

```{r, results='asis'}
hook(pbinom(.K, .N, .P), subsupscript(sum(dbinom(i, .N, .P)), i=0L, .K))
mathout(term)
```

@@

@@

For asynchronous stimuli, Schwarz [-@schwarz1994; Eq. 10] derived the expected
first-passage time as a function of the stimulus onset asyncrony \(\tau\),

>>>>

---
title: "mathml: Translate R expressions to MathML and LaTeX"
date: "2023-01-28"
abstract: >
  This R\ package translates R\ objects to suitable elements in MathML or LaTeX,
  thereby allowing for a pretty mathematical representation of R\ objects and
  functions in data analyses, scientific reports and interactive web content. In
  the RMarkdown document rendering language, R\ code and mathematical content
  already exist side-by-side. The present package allows to use the same
  R\ objects for both data analysis and typesetting in documents or web content.
  This tightens the link between the statistical analysis and its verbal
  description or symbolic representation, which is another step towards
  reproducible science. User-defined hooks allow to extend the package by
  mapping specific variables or functions to new MathML and LaTeX entities. A
  number of examples and a case study illustrate the use of the package in a
  scientific report.
draft: true
author:  
  - name: Matthias Gondan
    affiliation: Universität Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/gondan-rochon/index.html.en
    orcid: 0000-0001-9974-0057
    email:  Matthias.Gondan-Rochon@uibk.ac.at
  - name: Irene Alfarone
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/alfarone/index.html.en
    email: Irene.Alfarone@uibk.ac.at
    orcid: 0000-0002-8409-8900
    affiliation: Universität Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
type: package
output: 
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: bibliography.bib
preamble:
- \usepackage{cancel}
- \usepackage{amsmath}
Vignette: >
  %\VignetteIndexEntry{mathml: Translate R expressions to MathML and LaTeX}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# linear models: DepVar ~ X0 + X1

@@

```{r, results='asis'}
term <- quote(1 + -2L + a + abc + "a" + phi + Phi + varphi + roof(b)[i, j]^2L)
mathout(term)

@@

To avoid name clashes with package \pkg{stats}, `roof()` is used to put a hat
on a symbol (see next section for further decorations). Note that an 
R\ function `roof()` does not exist in base R, it is provided by the package
for convenience and points to the identity function.

## Decorations

The package offers some support for different fonts as well as accents and
boxes etc. Internally, these decorations are implemented as identity functions,
so they can be introduced into R expressions without side-effects.

```{r, results='asis'}
term <- quote(bold(b[x, 5L]) + bold(b[italic(x)]) + italic(ab) + italic(42L))
mathout(term)

term <- quote(tilde(a) + mean(X) + box(c) + cancel(d) + phantom(e) + prime(f))
mathout(term)
```

Note that the font styles only affect the display of identifiers, whereas
numbers, character strings etc. are left untouched.

@@

Indices in square brackets are rendered as subscripts, powers are rendered as
superscript. Moreover, \pkg{mathml} defines the
functions `sum_over(x, from, to)`, and `prod_over(x, from, to)` that simply
return their first argument. The other two arguments serve as
decorations (_to_ is optional), for example, for summation and product signs.

@@

term <- quote(log(prod_over(L[i], i==1L, N)) <- sum_over(log(L[i]), i==1L, N))
mathout(term)
```

@@

```{r, results='asis'}
hook(pbinom(.K, .N, .P), sum_over(dbinom(i, .N, .P), i=0L, .K))
mathout(term)
```

@@

```{r, results='asis'}
hook(successes, k)
hook(quote(Ntotal), quote(N), quote=FALSE)
hook(prob, pi)

term <- quote(dbinom(successes, Ntotal, prob))
mathout(term)
```

@@

For asynchronous stimuli, Schwarz [-@schwarz1994] derived the expected
first-passage time as a function of the stimulus onset asyncrony \(\tau\),
